---
title: "3.- TCGA_BRCA_DNA_methylation_DMR"
author: "Nuria Moragas"
date: "July 2022"
output:
  html_document:
---

In this doc it have been anilized methylation profile from TCGA_BRCA.


  * - 3.1.- Work environment preparation
  * - 3.2.- Methylation Data Download
  * - 3.3.- Row data procesing and quality control
  * - 3.4.- Differential methylation analysis - CpG
  * - 3.5.- Visualization - CpG
  * - 3.6.- CpGs POSITION 
  

*WEBS*

https://bioconductor.org/packages/devel/bioc/vignettes/TCGAbiolinks/inst/doc/analysis.html
http://openaccess.uoc.edu/webapps/o2/bitstream/10609/82726/6/mparragrandeTFM0618memoria.pdf 
https://www.costalab.org/wp-content/uploads/2020/11/R_class_D3.html#2_TCGA_data
https://bioconductor.org/packages/devel/bioc/vignettes/TCGAbiolinks/inst/doc/download_prepare.html
http://bioconductor.org/packages/release/workflows/vignettes/TCGAWorkflow/inst/doc/TCGAWorkflow.html
   
# 3.1 -Work environment preparation

This part include:
  - 3.1.1 - Directory Definition
  - 3.1.2 - Library Call
  - 3.1.3 - Color Palette
  


## 3.1.1 - Directory Definition

```{r message=FALSE, warning=FALSE}

#___________________Directori definition____________________#

setwd("./Scrips_R/3.- TCGA_BRCA_DNA_methylation")

```

## 3.1.2 - Library Call:

```{r message=FALSE, warning=FALSE}
#_________________________Packages_______________________#

library(knitr)
library(dplyr)
library(minfi )

library(edgeR)
library(limma)

library(TCGAbiolinks)
library(MultiAssayExperiment)
library(SummarizedExperiment)

library(RColorBrewer)
library(ggplot2)
library(EnhancedVolcano)

```

## 3.1.3 - Color Palette:

```{r}
 #...................................COLORS........................

colors_Gr_Bw <- c( "#35978F", "#80CDC1", "#C7EAE5", "#FFFFFF" , "#F6E8C3", "#DFC27D", "#BF812D" ,"#8C510A")
colors_bambino<- c("#AB0302", "#871C0F" , "#912303", "#DB6B3E", "#D37C1B", "#704511", "#BFAF97", "#6A5224", "#CBBEA3", "#6D4C03", "#BA9951", "#8F7C50", "#B1AC76", "#A6A48A", "#858465", "#566048", "#8FB9C4", "#AACFD9", "#607C85", "#8EA3AD")

```

  

# 3.2.- Methylation Data Download


```{r eval=FALSE, message=FALSE, warning=FALSE}

#_________________________Loading_samples barcode_______________________#

barcode_TM_IDC <-readRDS(file = '0.-barcode_TM_IDC.rda')
barcode_TP_IDC = readRDS(file = "0.-barcode_TP_IDC.rda")
load("0.-barcode_TP_IDC.rda")


IDC_TM_TP_barcode <- c(barcode_TM_IDC, barcode_TP_IDC )

saveRDS(object = IDC_TM_TP_barcode,
        file = "0.- IDC_TM_TP_barcode",
        compress = FALSE)


#_________________________Download_Data_______________________#




#IMPORT METHYLATION TCGA_BRCA DATA: 

# SOme information:
TCGAbiolinks:::getProjectSummary("TCGA-BRCA") 


# Query platform Illumina HiSeq with a list of barcode 
#references GRCh37 (hg19) and GRCh36 (hg18).
query_MET <- GDCquery(
    project = "TCGA-BRCA", 
    barcode = IDC_TM_TP_barcode,
    data.category = "DNA methylation",
    data.type = "Methylation beta value",
    platform = "Illumina Human Methylation 450", # Whith my computer is not possible download 450k
    legacy = TRUE,
    sample.type = c("Primary Tumor", "Metastatic"))


#To visualize the query results in a more readable way, we can use the command getResults.
            brca_met = getResults(query_MET) # make results as table
            # head(lihc_res) # data of the first 6 patients.
            colnames(brca_met) # columns present in the table
            
            head(brca_met)
            summary(factor(brca_met$sample_type)) #Metastatic Primary Tumor 
                                                  #  2           314 
            
            brca_met%>%
  group_by(sample_type)%>%
  summarise(n=n())%>%
  mutate(Perc.= n/sum(n)*100)

            summary(factor(brca_met$data_type))
            summary(factor(brca_met$experimental_strategy))
            summary(factor(brca_met$data_format))
            summary(factor(brca_met$analysis_workflow_type ))
            
   

# Download a list of barcodes with platform IlluminaHiSeq_RNASeqV2
GDCdownload(query_MET, method = "api")


BRCA_methylation <- GDCprepare(query_MET)


        #Look insdie this data:
         # Obtain the RNA-seq count matrices and rowData to see gene mapping information. How many genes and how many samples are included there?

      dim(assay(BRCA_methylation))# [1] 27578   316

          #assay() Make a matrix
      head(assay(BRCA_methylation)[,1:10]) # expression of first 6 genes and first 10 samples
      #Defult assau: unstranded
      

      head(rowData(BRCA_methylation, use.names=TRUE))  # ensembl id and gene id of the first 6 genes.
      
# DNA Methylation levels:
      BRCA_methylation %>% assay %>% head %>% as.data.frame


# In order to have not to reapeat the previos operation (GDCprepare(query)), BRCA.Rnaseq.SE is sav as RDS file. Is id needed, load it.

saveRDS(object = BRCA_methylation,
        file = "3.-BRCA_Methylation_metastasi_450_primariTumor_select.rda",
        compress = FALSE)

saveRDS(object = BRCA_methylation,
        file = "3.-BRCA_Methylation_metastasi_450_primariTumor_select.rda_",
        compress = T)
```

```{r}
      #Loading data:
BRCA_methylation = readRDS(file = "3.-BRCA_Methylation_metastasi_450_primariTumor_select.rda")


BRCA_methylation%>%rowRanges%>%as.data.frame%>%head

#You can access the DNA methylation levels with assay.

BRCA_methylation%>% assay %>% head %>% as.data.frame
```

 
# 3.3 - Row data procesing and quality control

      * - 3.3.1.- Data Group checking
      * - 3.3.2.- Data Filtering: NA values and same beta-values Delation
      * - 3.3.3.- Quality control. Data exploration - visualization    
      
      

      
## 3.3.1.- Data Grup comprovations


```{r}
#________________________ Data Group checking ________________________#

#DATA:BRCA_Rnaseq - primary_diagnosis 1106
#Subtye selection:


table(BRCA_methylation@colData$sample_type) #M 4   PT 319
table(BRCA_methylation@colData$primary_diagnosis)
table(BRCA_methylation@colData$ajcc_pathologic_m) #M0 323
table(BRCA_methylation@colData$ajcc_pathologic_stage)
#Stage I  Stage IA  Stage IB  Stage II Stage IIA Stage IIB 
#       38        34         3         3       143       102 
table(BRCA_methylation@colData$primary_diagnosis) # IDC 323
table(BRCA_methylation@colData$gender)# female 323
dim(BRCA_methylation@colData) #[1] 323   85


```

## 3.3.2.- Data Filtering: NA values and same beta-values Delation
     
    * A.- remove probes with NA    
    * B.- remove probes that match to chromosome  X and Y (Not applay)
    * C.- remove SNPs overlapped probe
    * D .- SNPs with maf <= 0.05
    * E.- Removing probes that have been demonstrated to map to multiple places in the genome.
    
```{r}
as.data.frame(BRCA_methylation@rowRanges)
length(BRCA_methylation@rowRanges)#485577
```

```{r}
##___________ F iltering methylation data ______________#


#met matrix
BRCA_met_matrix <- as.data.frame(SummarizedExperiment::assay(BRCA_methylation))
head(BRCA_met_matrix)
dim(BRCA_met_matrix ) #[1] 485577    323


## A.- remove probes with NA
probe.na <- rowSums(is.na(BRCA_met_matrix ))

table(probe.na == 0)
 #FALSE   TRUE 
#109604 375973 

# chose those has not NA values in rows
probe <- probe.na[probe.na == 0]
BRCA_met_matrix <- BRCA_met_matrix[row.names(BRCA_met_matrix) %in% names(probe), ]
dim(BRCA_met_matrix)#[1] 375973    323 
rm(probe.na, probe)

# get the 450k annotation data
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)

## -- NOT APPLIED -- because there are only women. 
## B.  remove probes that match to chromosome  X and Y 
#keep <- !(row.names(BRCA_met_matrix ) %in% ann450k$Name[ann450k$chr %in% c("chrX","chrY")])
#table(keep)
#BRCA_met_matrix  <- BRCA_met_matrix [keep, ]
#rm(keep) 


## C. remove SNPs overlapped probe


table(is.na(ann450k$Probe_rs))
# probes without snp
no.snp.probe <- ann450k$Name[is.na(ann450k$Probe_rs)]
snp.probe <- ann450k[!is.na(ann450k$Probe_rs), ]



## D .- SNPs with maf <= 0.05
snp5.probe <- snp.probe$Name[snp.probe$Probe_maf <= 0.05]
# filtre BRCA_met_matrix
BRCA_met_matrix <- BRCA_met_matrix[row.names(BRCA_met_matrix) %in% c(no.snp.probe, snp5.probe), ]

#remove no-further needed dataset
rm(no.snp.probe, snp.probe, snp5.probe)


##E.- Removing probes that have been demonstrated to map to multiple places in the genome.
# list adapted from https://www.tandfonline.com/doi/full/10.4161/epi.23470

crs.reac <- read.csv("cross_reactive_probe.chen2013.csv")
crs.reac <- crs.reac$TargetID[-1]

# filtre met
BRCA_met_matrix <- BRCA_met_matrix[ -which(row.names(BRCA_met_matrix) %in% crs.reac), ]

```

```{r}
## converting beta values to m_values
bval <- BRCA_met_matrix
## m = log2(beta/1-beta)
mval <- t(apply(BRCA_met_matrix, 1, function(x) log2(x/(1-x))))

rm(BRCA_met_matrix)

head(bval)
head(mval)
```

```{r}
#______________saving/loading_____________________#
# save data sets
saveRDS(mval, file = "3.- mval.RDS", compress = FALSE)
saveRDS (bval, file = "3.- bval.RDS", compress = FALSE)
mval <- readRDS("3.- mval.RDS")
bval <- readRDS("3.- bval.RDS")


bval_matrix <- data.matrix(bval, rownames.force = NA)
mval_matrix <- data.matrix(mval, rownames.force = NA)

```



##  - 3.3.3.- Quality control. Data exploration - visualization


```{r}
#Row data
row_matrix <- data.matrix(BRCA_met_matrix, rownames.force = NA)
hist(row_matrix, col = colors_Gr_Bw, breaks = 50 )

```

```{r}

##b and M values distribution:
#Noramlized - filt
bval_matrix <- data.matrix(bval, rownames.force = NA)
hist(bval_matrix, col = colors_Gr_Bw, breaks = 50 )

```

```{r}
#Density plot
library(minfi )
par(mfrow=c(1,2))

png(file="Plots_results/3.0.- Density_plot_bval_norm.png" #,  width =  500 , height = 400 , units = "px" 
    )
densityPlot(bval_matrix,legend = TRUE, main = "BRCA beta-values",sampGroups = BRCA_methylation$sample_type, pal = c("#8C510A", "#35978F"))
dev.off()

```


```{r}

# MDS plots to look at largest sources of variation - SI FUNCIONA
ggpubr::show_point_shapes()#per veure les formes

par(mfrow=c(1,2))
#pal <- brewer.pal(8,"Dark2")
pal <- c("#AB0302", "#108372")


png(file="Plots_results/3.0.- PCA_bval_norm_TM_TP.png" #,  width =  500 , height = 400 , units = "px" 
    )
plotMDS(bval_matrix, top=5000, 
        pch = 19, #Forma 
        gene.selection="common", 
        col=pal[factor(BRCA_methylation$sample_type)])
legend("topright", legend=levels(factor(BRCA_methylation$sample_type)), text.col=pal, bg="white", cex=1)

dev.off()
```


```{r}
#PCA subtype

pal <- c("#DFC27D", "#AB0302", "#108372", "#8C510A","tomato" )


png(file="Plots_results/3.0.- PCA_bval_norm_PAM50.png" #,  width =  500 , height = 400 , units = "px" 
    )
plotMDS(bval_matrix, top=5000, 
        pch = 19, #Forma 
        gene.selection="common", 
        col=pal[factor(BRCA_methylation$paper_BRCA_Subtype_PAM50 )])
legend("topright", legend=levels(factor(BRCA_methylation$paper_BRCA_Subtype_PAM50)), text.col=pal, bg="white", cex=1)

dev.off()


```

### - DNA methylation mean

```{r}

df <- data.frame(
  "Sample.mean" = colMeans((bval), na.rm = TRUE),
  "groups" = BRCA_methylation$sample_type)

png(file="Plots_results/3.0.- PCA_bval_mean_TM_vs_TP.png" #,  width =  500 , height = 400 , units = "px" 
    )

library(ggpubr)
ggpubr::ggboxplot(
  data = df,
  y = "Sample.mean",
  x = "groups",
  color = "groups",
  add = "jitter",
  fill = "white",
  palette = c("#AB0302", "#108372"),
  ylab = expression(paste("Mean DNA methylation (", beta, "-values)")),
  xlab = ""
) + stat_compare_means() 

dev.off()

```



# 3.4.- Differential methylation analysis

__________________ DIFERENTION MTHYLATION ANALYSIS CpGs _____________

------- Different methylated CpGs (DMC) or Probe-wise differential methylation analysis ------ 

```{r}
# call clinical data:
clinical_data=data.frame(colData(BRCA_methylation))
head(clinical_data)
```

```{r}
#The clinical data is saved:
saveRDS(object = clinical_data,
        file = "3.- clinical_data_meta.RDS",
        compress = FALSE)

clinical_data = readRDS(file = "3.- clinical_data_meta.RDS")

```



```{r}
#Group of interest:
group=factor(clinical_data$shortLetterCode)
table(group)
group=relevel(group, ref = "TP")
table(group)


#Matrix design
design=model.matrix(~0+group, data = clinical_data) #group defined before
head(design)

#Re-name matrix
colnames(design)<- gsub("group", "", colnames(design)) # group name delate
rownames(design) <- colnames(bval)
head(design)

#fit the linear model
fit <- lmFit(mval_matrix, design)

#CONTRAST MATRIX creation
contr.matrix <- makeContrasts(
  TMvsTP = TM-TP,
  levels= colnames(design)
)
contr.matrix


#Fit the contrast
cfit <- contrasts.fit(fit, contr.matrix)
fit2 <- eBayes(cfit)
plotSA(fit2)

```


```{r}
#TOPTABLE DMC
# get the table of results 
ann450kSub <- ann450k[match(rownames(mval_matrix),ann450k$Name),
                      c(1:4,11:19,24:ncol(ann450k))]#Aquets són les columnes que interesen afegir. 


#as.data.frame(ann450k)
#All methilated proves
All_meth <- topTable(fit2, #ANTIGAMENT = DMCs
                 num=Inf, 
                 sort.by = "p",
                 coef=1, #saved as data.frame
                 #p.value = 0.05,
                 genelist=ann450kSub)

dim(All_meth) #[1] 333315     29

#Gene column #UCSC_RefGene_Name en alguns casos te escrit varis cops el nom del gen
All_meth$Gene <-sapply(sapply(strsplit(All_meth$UCSC_RefGene_Name, ";"), unique), function(x) ifelse(length(x)>1, paste(x[1], x[2], sep = "_"), x[1])) 

All_meth$CpgGene <- paste(All_meth$Row.names, All_meth$Gene)


#Significant DMCs:
#Select significat DMCs, hyper and hypomethylated
library(tidyverse) #colecio de package
DMCs_sig <- All_meth %>% filter(All_meth$P.Value < 0.05) # Idem calculat anteriorment, 
dim(DMCs_sig) # [1] 34128    31

saveRDS(DMCs_sig,file = "3.- BRCA_DMG_TM_sig_genes.RDS")
write.table(DMCs_sig, file="Plots_results/3.- BRCA_DMCs_sig_CpG.csv", sep=",", row.names=T)

```


```{r}
saveRDS(All_meth, file = "3.- All_meth.RDS", compress = FALSE)
All_meth <- readRDS("All_meth.RDS")

```

```{r}
# ____________________  UP ______________________________ 
## HYPER - pvalue
DMCs_up <- DMCs_sig %>% filter(logFC>0)
DMCs_up <- DMCs_up[order(DMCs_up$P.Value),]

dim(DMCs_up) # [1] 3326   31

top5pval_up <- DMCs_up[1:5,]
top5pval_up$name <- gsub( " ", "\n", top5pval_up$UCSC_RefGene_Name)


## HYPER - FC
DMCs_up_FC <- DMCs_sig %>% filter(logFC>0)%>% arrange(desc(logFC))
dim(DMCs_up_FC) #[1] 3326   31

top5FC_up <- DMCs_up_FC[1:5,]
top5FC_up$name <- gsub( " ", "\n", top5FC_up$CpgGene)


# ______________________________  DOWN _______________________________
## HYPO - pvalue
DMCs_down <- DMCs_sig %>% filter(logFC<0)
DMCs_down <- DMCs_down[order(DMCs_down$P.Value),]

dim(DMCs_down)#[1]  [1] 30802    31

top5pval_down <- DMCs_down[1:5,]
top5pval_down$name <- gsub( " ", "\n", top5pval_down$CpgGene)

## HYPO - FC

DMCs_down_FC <- DMCs_sig %>% filter(logFC<0)%>% arrange(logFC)
top5FC_down <- DMCs_down_FC[1:5,]
top5FC_down$name <- gsub( " ", "\n", top5FC_down$CpgGene)

## COmbinacio
top10_up <- rbind(top5pval_up, top5FC_up)
dim(top10_up)
top10_down <- rbind(top5pval_down, top5FC_down)
dim(top10_down)
```



# - 3.5.- Visualization - CpG


VISUALIZATION
- BoxPlot probes (DMCs)

```{r}
#Visualization 


#OPCIO 1
# plot the top 10 most significantly differentially methylated CpGs 

par(mfrow=c(2,5))
png(file="Plots_results/3.1.- Top_10_diff_CpGs_MP.png" #,  width =  500 , height = 400 , units = "px" 
    )
sapply(rownames(DMCs_sig)[1:10], function(cpg){
  plotCpg(bval, cpg=cpg, pheno= clinical_data$shortLetterCode, ylab = "Beta values")
})
dev.off()
```


```{r}

#OPCIO 2


# plot the top 4 most significantly differentially methylated CpGs 
png(file="Plots_results/3.1.- top few differentially methylated CpGs.png" #,  width =  500 , height = 400 , units = "px" 
    )
par(mfrow=c(2,2))
sapply(rownames(DMCs_sig)[1:4], function(cpg){
  plotCpg(bval_matrix, cpg=cpg, pheno=clinical_data$shortLetterCode, ylab = "Beta values")
})

dev.off()
```


- VOLCANO PLOT


```{r}

#-------------------------------- VOLCANO PLOT -------------------------

#OPCIO 1 - IGUAL QUE RNA seq 

#Gene column #UCSC_RefGene_Name en alguns casos te escrit varis cops el nom del gen
All_meth$Gene <-sapply(sapply(strsplit(All_meth$UCSC_RefGene_Name, ";"), unique), function(x) ifelse(length(x)>1, paste(x[1], x[2], sep = "_"), x[1]))



png(file="Plots_results/3.1.- Volcano plot DMR log.png",  width =  500 , height = 400 , units = "px" 
    )

EnhancedVolcano(All_meth,
  lab = All_meth$Gene, # rownames(DMPs), #dfinit mes endavant Gene
  x = "logFC",
  y = "adj.P.Val",
  pCutoff = 10e-3,
  FCcutoff = 0.5,
  xlim = c(-3, 3),
  ylim = c(0, -log10(10e-12)),
  pointSize = 1.0,
  labSize = 4.0,
  title = '',
  subtitle = '',
  caption = 'FC cutoff, 0.5; p-value cutoff, 10e-3',
  legendPosition = "right",
  legendLabSize = 10,
  col = c('grey30', "#35978F", 'gray81', "#8C510A"),
  colAlpha = 0.5,
  drawConnectors = FALSE,
  hline = c(10e-8),
  widthConnectors = 0.5,
  max.overlaps = 15)

dev.off()
```


- MANHATTAN PLOT

```{r}
# https://github.com/YinLiLin/CMplot

library("CMplot")
?CMplot

#Slecionem columnes interes (nom, chr, pos i p-value)
DMCs_manhattan<- DMCs_sig[,c(1,2,4,27)] 
#rownames remove
rownames(DMCs_manhattan) <- NULL
#Rearrange colums
DMCs_manhattan<- DMCs_manhattan[,c(3,1,2,4)]
#Rename chr
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr1"] <- "1"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr2"] <- "2"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr3"] <- "3"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr4"] <- "4"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr5"] <- "5"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr6"] <- "6"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr7"] <- "7"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr8"] <- "8"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr9"] <- "9"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr10"] <- "10"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr11"] <- "11"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr12"] <- "12"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr13"] <- "13"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr14"] <- "14"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr15"] <- "15"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr16"] <- "16"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr17"] <- "17"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr18"] <- "18"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr19"] <- "19"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr20"] <- "20"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr21"] <- "21"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chr22"] <- "22"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chrX"] <- "X"
DMCs_manhattan["chr"][DMCs_manhattan["chr"] == "chrY"] <- "Y"

#sort column chr numerically
DMCs_manhattan <- DMCs_manhattan[order(DMCs_manhattan$chr),] 

DMCs_manhattan


png(file="Plots_results/3.2.- Manhattan_plot_DMR.png"#,  width =  500 , height = 400 , units = "px" 
    )

CMplot(DMCs_manhattan, 
       plot.type="m", #Rectangular Manhattan Plot, if "c" = Circular 
       ylim= NULL, # if the ylim is setted, then CMplot will only plot the points among this interval,
       
       #Significant level:
       
        LOG10= T, #change p-value to log10(p-value)
       threshold= c(10e-6),#sig level, Coluld be 0 or NULL, defult = 0.01
       threshold.lty=c(1), #line type. 1 = continuos 2 = discontinous
        threshold.lwd=c(1),#Threshold line width 
       threshold.col=c("black"), #line colour
       
       #Highlight the point which are significant (thereshold):
       amplify= F,
       signal.cex=c(0.8,0.8), #size of sig. points
       signal.col=c("red","green"), #Color
       signal.pch=c(20,20), #shape sig. points
       
       
       #Chromosome density on the bottom
       #chr.den.col=c("darkgreen", "yellow", "red"),
       
       #Points Colours/size 
       cex = 0.8,#size of the points
       pch = 20, #type of the points
       band = 2,# space between chromosomes 
       bin.size=1e6,
       col = c("#80CDC1","#DFC27D"),#point colour
       
       chr.border=TRUE, #Vertical line between chr
       
      
       #Legend, title and size numbers etc
       cex.axis = 0.5,#Size numbers x axis
       main = "Manhattan Plot DMCs", #Title
       main.cex = , #size of the title
       #ylab = expression(-log[2](italic(p))), 
       
       #Output File creation:
       file.output=F,
       file="jpg",
       memo="", #add a character to the output file name
       dpi=300, #the picture resolution
       verbose=TRUE,
       width=14,height=6)

dev.off()

```

```{r}
#Chr precentage - statistics
table(DMCs_sig$chr)

df<- DMCs_sig%>%
  group_by(chr)%>%
  summarise(n=n())%>%
  mutate(Perc.= n/sum(n)*100)

df2 <- df[order(df$Perc.,decreasing=TRUE),]
df2

```



# - 3.6.- CpGs POSITION 

* - 3.6.1.- RELATION to GENE 
* - 3.6.2.- RELATION to ISLAND 


## 3.6.1.- RELATION to GENE 

---------  (DMC) CpGs POSITION / RELATION to GENE  ----------- 

    - Regulatory (TSS200, TSS1500, 1StExon, 5'UTR)
    - Body Regions
    - 3'UTR Regions
    
    
    
```{r}
#
#Variable: UCSC_RefGene_Group - All vs DMCs_Hyper vs DMCs_Hypo
as.data.frame(table(DMCs_sig$UCSC_RefGene_Group))
#(c) Percentage of CpGs according to relation to genes in DMPs (Vacc-BSL) and in 450K array (All).  


#All CpG - All_meth
#CpG sig - DMCs_sig


#Colour definition:
pal <- brewer.pal(8,"BrBG") # 

#1.- UCSC Gene Group annotation modification (DMCs group and All CpG as control):

DMCs_up$Gene_Group <- sapply(sapply(strsplit(DMCs_up$UCSC_RefGene_Group, ";"), unique), function(x)x[1])
DMCs_down$Gene_Group <- sapply(sapply(strsplit(DMCs_down$UCSC_RefGene_Group, ";"), unique), function(x)x[1])
All_meth$Gene_Group <- sapply(sapply(strsplit(All_meth$UCSC_RefGene_Group, ";"), unique), function(x)x[1])

```


```{r}

#2.- Ratio calculation : CpG position / ralation to Gene

## 2.1 .- For DMCs_up:

Relation_DMCs_up <- DMCs_up %>%
  group_by(Gene_Group) %>%
  summarise(counts = n())
Relation_DMCs_up

#Relation_DMCs$x <- rep("Relation to nearest gene", nrow(Relation_DMCs))
#Relation_DMCs_up$Gene_Group  <- factor(Relation_DMCs$Gene_Group, levels = c("TSS200", "TSS1500", "1stExon", "5'UTR", "3'UTR", "Body"))

png(file="Plots_results/3.3.- CpG position_ralation to Gene_up.png"#,  width =  500 , height = 400 , units = "px" 
    )

ggplot(Relation_DMCs_up, aes(fill=Gene_Group, y=counts, x = "Relation to nearest gene")) + 
    geom_bar(position="fill", stat="identity") + 
  theme_classic()+scale_fill_manual(values = pal)+
  guides(fill=guide_legend(title="Genomic region"))

dev.off()
```

```{r}
## 2.2 .- For DMCs_down:

Relation_DMCs_down <- DMCs_down %>%
  group_by(Gene_Group) %>%
  summarise(counts = n())
Relation_DMCs_down

png(file="Plots_results/3.3.- CpG position_ralation to Gene_down.png"#,  width =  500 , height = 400 , units = "px" 
    )

ggplot(Relation_DMCs_down, aes(fill=Gene_Group, 
       y=counts, x = "Relation to nearest gene")) + 
      geom_bar(position="fill", stat="identity") +
      theme_classic()+scale_fill_manual(values = pal)+  guides(fill=guide_legend(title="Genomic region"))

dev.off
```

```{r}
## 2.3 .- For All_meth:


Relation_All_meth <- All_meth %>%
  group_by(Gene_Group) %>%
  summarise(counts = n())
Relation_All_meth

png(file="Plots_results/3.3.- CpG position_ralation to Gene_ALL.png"#,  width =  500 , height = 400 , units = "px" 
    )

ggplot(Relation_All_meth, aes(fill=Gene_Group, y=counts, x = "Relation to nearest gene")) + 
    geom_bar(position="fill", stat="identity") + 
  theme_classic()+scale_fill_manual(values = pal)

dev.off
```


```{r}
# PLOT ALL TOGETHER
#3.- Class varaiable Cration: All, Hyper, Hypo
Relation_All_meth$class <- rep("All", nrow(Relation_All_meth))
Relation_DMCs_up$class <- rep("Hyper", nrow(Relation_DMCs_up))
Relation_DMCs_down$class <- rep("Hypo", nrow(Relation_DMCs_down))

#4.- % calculation
Relation_All_meth$perc <- 100*(Relation_All_meth$counts/sum(Relation_All_meth$counts))
Relation_DMCs_up$perc <- 100*(Relation_DMCs_up$counts/sum(Relation_DMCs_up$counts))
Relation_DMCs_down$perc <- 100*(Relation_DMCs_down$counts/sum(Relation_DMCs_down$counts))

#5.- Combine all data frame
Gene_group_df <- rbind(Relation_All_meth, Relation_DMCs_up, Relation_DMCs_down)

#Change NA to IGR. Intergenic Gene Regions
Gene_group_df$Gene_Group[is.na(Gene_group_df$Gene_Group)] <- "IGR"
Gene_group_df



# 6.- Plot

Gene_group_df$p1 <- with(Gene_group_df, paste(formatC(perc, digits=2
              , format="fg"),"%",sep=""))

png(file="Plots_results/3.3.- CpG position_ralation to Gene_COMBINATION.png"#,  width =  500 , height = 400 , units = "px" 
    )

ggplot(Gene_group_df, 
       aes(fill=Gene_Group, y=perc, x = class)) + 
    geom_bar(position="fill", stat="identity")+
  
  theme(axis.text.x  = element_text(colour = "black", size = 11), 
  axis.text.y = element_text(colour = "black", size = 11), 
  legend.text = element_text(size = 10,  colour ="black"), 
  legend.title = element_text(size = 12), 
  panel.background = element_blank(), axis.line = element_line(colour = "black", size = 1), 
  legend.position = "right") +
  
  scale_fill_manual(values = pal) + 
  
  #% definition:
  scale_y_continuous(labels = scales:: percent_format(accuracy = 1))+
  
  labs(x = "", y = "% of CpG positions / Relative to Gene")+
  geom_text(aes(label=p1),position = position_fill(vjust = 0.5))

dev.off()


#7. Statistics
library(ggpubr)
Gene_group_df

p_st <- compare_means(perc ~ class, data = Gene_group_df, 
              group.by = "Gene_Group", paired = TRUE)

write.table(p_st, file="Plots_results/3.3.- CpG position_ralation to Gene_COMBINATION_statistics.csv", sep=",", row.names=F)
 
```

## 3.6.2.- RELATION to ISLAND 
---------  (DMC) CpGs POSITION / RELATION to ISLAND  ----------- 

    - CpG Island (CGI)
    - CpG Island Shores (2kb from CGI)
    - CpG Island Shelve (4kb from CGI)
    - Open See Regions
    
```{r}

#(d) Percentage of CpGs according to relation to Island in DMPs (Vacc-BSL) and in 450K array (All).
#Variable: Relation_to_Island -  All vs DMCs_Hyper vs DMCs_Hypo

as.data.frame(table(DMCs_sig$Relation_to_Island))
#(c) Percentage of CpGs according to relation to genes in DMPs (Vacc-BSL) and in 450K array (All).  


#Colour definition:
pal <- brewer.pal(8,"BrBG") # 
```

```{r}
#1.- UCSC Gene Group annotation modification (DMCs group and All CpG as control):
## No es necessari modificar anotacio

#2.- Ratio calculation : CpG position / ralation to Gene

## 2.1 .- For DMCs_up:

Relation_DMCs_up_island <- DMCs_up %>%
  group_by(Relation_to_Island) %>%
  summarise(counts = n())
Relation_DMCs_up_island

ggplot(Relation_DMCs_up_island, aes( fill=
Relation_to_Island,y=counts, x = "Hyper")) + 
    geom_bar(position="fill", stat="identity") + 
  theme_classic()+scale_fill_manual(values = pal)
```

```{r}
## 2.2 .- For DMCs_down:

Relation_DMCs_down_island <- DMCs_down %>%
  group_by(Relation_to_Island) %>%
  summarise(counts = n())
Relation_DMCs_down

ggplot(Relation_DMCs_down_island, aes(fill=
Relation_to_Island,y=counts, x = "Hypo")) + 
      geom_bar(position="fill", stat="identity") +
      theme_classic()+scale_fill_manual(values = pal)
```

```{r}
## 2.3 .- For All_meth:

Relation_All_meth_island <- All_meth %>%
  group_by(Relation_to_Island) %>%
  summarise(counts = n())
Relation_All_meth

ggplot(Relation_All_meth_island, aes(fill=Relation_to_Island, y=counts, x = "Relation to nearest gene")) + 
    geom_bar(position="fill", stat="identity") + 
  theme_classic()+scale_fill_manual(values = pal)
```

```{r}
#COMPINATION ALL
#3.- Class varaiable Cration: All, Hyper, Hypo
Relation_All_meth_island$class <- rep("All", nrow(Relation_All_meth_island))
Relation_DMCs_up_island$class <- rep("Hyper", nrow(Relation_DMCs_up_island))
Relation_DMCs_down_island$class <- rep("Hypo", nrow(Relation_DMCs_down_island))

#4.- % calculation
Relation_All_meth_island$perc <- 100*(Relation_All_meth_island$counts/sum(Relation_All_meth_island$counts))
Relation_DMCs_up_island$perc <- 100*(Relation_DMCs_up_island$counts/sum(Relation_DMCs_up_island$counts))
Relation_DMCs_down_island$perc <- 100*(Relation_DMCs_down_island$counts/sum(Relation_DMCs_down_island$counts))

#5.- Combine all data frame
Gene_group_df_island <- rbind(Relation_All_meth_island, Relation_DMCs_down_island, Relation_DMCs_up_island)


# 6.- Plot

Gene_group_df_island$p2 <- with(Gene_group_df_island, paste(formatC(perc, digits=2
              , format="fg"),"%",sep=""))
  Gene_group_df_island
  
  
png(file="Plots_results/3.3.- CpG position Relation_to_Island COMBINATION.png"#,  width =  500 , height = 400 , units = "px" 
    )  
  
ggplot(Gene_group_df_island, 
       aes(fill=Relation_to_Island, y=perc, x = class)) + 
    geom_bar(position="fill", stat="identity")+
  
  theme(axis.text.x  = element_text(colour = "black", size = 11), 
  axis.text.y = element_text(colour = "black", size = 11), 
  legend.text = element_text(size = 10,  colour ="black"), 
  legend.title = element_text(size = 12), 
  panel.background = element_blank(), axis.line = element_line(colour = "black", size = 1), 
  legend.position = "right") +
  
  scale_fill_manual(values = pal) + 
  
  #% definition:
scale_y_continuous(labels = scales:: percent_format(accuracy = 1))+
labs(x = "", y = "% of CpG positions / Relative to Island")+
geom_text(aes(label=p2),position = position_fill(vjust = 0.5))
  

#7 Statistics
p_st_2 <- compare_means(perc ~ class, data = Gene_group_df_island, 
              group.by = "Relation_to_Island", paired = TRUE)

write.table(p_st, file="Plots_results/3.3.- CpG position Relation_to_Island COMBINATION_statistics.csv", sep=",", row.names=F)

```
